---
title: "Data Analysis Assignment"
output: html_notebook
---

# This part is to stimulate the data 
```{r}
# load required data 
library(ggplot2)
library(dplyr)
library(tidyr)

# stimulate the data 

set.seed(20)

# create the total number of participants for this data frame 
nsubs = 34 # participants 

# example with 2 subpop
nGroups = 2
nSubsGroup = 18

# Condition
nEnvi = 2 # environment conditions 
EnviLabels <- c("Urban", "Rural")
EnvPerParticipant = 17


df <- data.frame(Participant=rep(0,nsubs),
                 Environment=rep("NA",nsubs),
                 ReactionTimeOverall=rep(0,nsubs),
                 ReactionTimeIncorOverall=rep(0,nsubs),
                 ReactionTime90=rep(0,nsubs),
                 IncoReactionTime90=rep(0,nsubs),
                 ReactionTime45=rep(0,nsubs),
                 IncoReactionTime45=rep(0,nsubs),
                 AccuracyOverall=rep(0,nsubs),
                 Accuracy90=rep(0,nsubs),
                 Accuracy45=rep(0,nsubs),
                 Age=rep(0,nsubs),
                 stringsAsFactors = TRUE)

head(df)

# Participant Data 
df$Participant <- seq(1,nsubs)

# Age of Participants 
df$Age = sample(18:50, nsubs, replace = TRUE)

# Environment Data 
en1 <- rep(EnviLabels[1], times = nsubs/2)
en2 <- rep(EnviLabels[2], times = nsubs/2)
envs <- c(en1,en2)
envs_shuffle <- sample(envs)
df$Environment <- rep(envs_shuffle)


df$ReactionTimeOverall = rnorm(nsubs, mean = 700, sd = 100)
df$ReactionTime90 = rnorm(nsubs, mean = 700, sd = 100)
df$ReactionTimeIncorOverall = rnorm(nsubs, mean = 800, sd = 100)
df$IncoReactionTime90 = rnorm(nsubs, mean = 750, sd = 100)
df$ReactionTime45 = rnorm(nsubs, mean = 720, sd = 100)
df$IncoReactionTime45 = rnorm(nsubs, mean = 770, sd = 100)

# Adjusting the reaction times for 90 degrees based on the environment condition
df$ReactionTime90[df$Environment == "Urban"] = df$ReactionTime90[df$Environment == "Urban"] - 200  # Urban with 90 degrees is faster
df$ReactionTime90[df$Environment == "Rural"] = df$ReactionTime90[df$Environment == "Rural"] + 200  # Rural with 90 degrees is slower

# Adjusting the incorrect reaction times for 90 degrees based on environment condition
df$IncoReactionTime90[df$Environment == "Urban"] = df$IncoReactionTime90[df$Environment == "Urban"] - 150  # Urban incorrect RT for 90 degrees is faster
df$IncoReactionTime90[df$Environment == "Rural"] = df$IncoReactionTime90[df$Environment == "Rural"] + 175  # Rural incorrect RT for 90 degrees is slower

# Adjusting the reaction times for 45 degrees based on the environment condition
df$ReactionTime45[df$Environment == "Urban"] = df$ReactionTime45[df$Environment == "Urban"] - 200  # Urban with 45 degrees is faster
df$ReactionTime45[df$Environment == "Rural"] = df$ReactionTime45[df$Environment == "Rural"] + 200  # Rural with 45 degrees is slower

# Adjusting the incorrect reaction times for 45 degrees based on environment condition
df$IncoReactionTime45[df$Environment == "Urban"] = df$IncoReactionTime45[df$Environment == "Urban"] - 200  # Urban incoherent RT for 45 degrees is faster
df$IncoReactionTime45[df$Environment == "Rural"] = df$IncoReactionTime45[df$Environment == "Rural"] + 200  # Rural incoherent RT for 45 degrees is slower


# This will give accuracy values as percentages between the given values 
df$AccuracyOverall = runif(nsubs, min=60, max=100)  
df$Accuracy90 = runif(nsubs, min=80, max=100)  
df$Accuracy45 = runif(nsubs, min=60, max=100)  

# Adjust accuracy based on environment condition  
df$AccuracyOverall[df$Environment == "Urban"] = df$AccuracyOverall[df$Environment == "Urban"] + 5  # Urban is 5% more accurate  
df$Accuracy90[df$Environment == "Urban"] = df$Accuracy90[df$Environment == "Urban"] + 5  
df$Accuracy45[df$Environment == "Urban"] = df$Accuracy45[df$Environment == "Urban"] + 5  

df$AccuracyOverall[df$Environment == "Rural"] = df$AccuracyOverall[df$Environment == "Rural"] - 5  # Rural is 5% less accurate  
df$Accuracy90[df$Environment == "Rural"] = df$Accuracy90[df$Environment == "Rural"] - 5  
df$Accuracy45[df$Environment == "Rural"] = df$Accuracy45[df$Environment == "Rural"] - 5  

# This is to make sure that accuracy stays between 0% and 100%
df$AccuracyOverall <- pmax(pmin(df$AccuracyOverall, 100), 0)  
df$Accuracy90 <- pmax(pmin(df$Accuracy90, 100), 0)  
df$Accuracy45 <- pmax(pmin(df$Accuracy45, 100), 0)  



```


# Data Cleaning 
```{r}

head(df) # Just to make sure that the data looks good 
summary(df) #can see all the basic information including the mean median, etc. 

 

```


# Data exploration 
```{r}
# explore the summary 

# THE MEAN VALUES 
meanRT <- mean(df$ReactionTimeOverall) # Mean for the reaction time for the correct trials 
meanIncRT <- mean(df$ReactionTimeIncorOverall) # Mean for the reaction time for the incorrect trials 
meanRT_90 <- mean(df$ReactionTime90) # Mean for the correct 90 degree rotations --> reaction time 
meanRT_45 <- mean(df$ReactionTime45) # Mean for the correct 45 degree rotations --> reaction time 
meanIncRT_90 <- mean(df$IncoReactionTime90) # Mean for the incorrect 90 degree rotations --> reaction time 
meanIncRT_45 <- mean(df$IncoReactionTime45) # Mean for the incorrect 45 degree rotations --> reaction time 
meanAcc <- mean(df$AccuracyOverall) # Mean for accuracy overall 
meanAcc_90 <- mean(df$Accuracy90) # Mean for percent accuracy in 90 degree trials 
meanAcc_45 <- mean(df$Accuracy45) # Mean for percent accuracy in 45 degree trials 
meanAge <- mean(df$Age) # Mean for age 

print(meanRT)
print(meanIncRT)
print(meanRT_90)
print(meanRT_45) 
print(meanIncRT_90) 
print(meanIncRT_45)
print(meanAcc)
print(meanAcc_90)
print(meanAcc_45)
print(meanAge)



# THE STANDARD DEVIATION VALUES 
sdRT <- sd(df$ReactionTimeOverall) # standard deviation for the reaction time for the correct trials 
sdIncRT <- sd(df$ReactionTimeIncorOverall) # standard deviation for the reaction time for the incorrect trials 
sdRT_90 <- sd(df$ReactionTime90) # standard deviation for the correct 90 degree rotations
sdRT_45 <- sd(df$ReactionTime45) # standard deviation for the correct 45 degree rotations
sdIncRT_90 <- sd(df$IncoReactionTime90) # standard deviation for the incorrect 90 degree rotations 
sdIncRT_45 <- sd(df$IncoReactionTime45) # standard deviation for the incorrect 45 degree rotations
sdAcc <- sd(df$AccuracyOverall) # standard deviation for accuracy overall 
sdAcc_90 <- sd(df$Accuracy90) # standard deviation for percent accuracy in 90 degree trials 
sdAcc_45 <- sd(df$Accuracy45) # standard deviation for percent accuracy in 45 degree trials 
sdAge <- sd(df$Age) # standard deviation for age 

print(sdRT)
print(sdIncRT)
print(sdRT_90)
print(sdRT_45) 
print(sdIncRT_90) 
print(sdIncRT_45)
print(sdAcc)
print(sdAcc_90)
print(sdAcc_45)
print(sdAge)


```

# Checking for Normality: 
## p-value < 0.05: Data are not normally distributed.
## p-value ≥ 0.05: Data are likely normally distributed
## If data is not normal, then alternative testing is required. 

```{r}

# Check normality using histograms, and Shapiro-Wilk test

# Reaction Time Overall Histogram
ggplot(df, aes(x = ReactionTimeOverall)) + 
  geom_histogram(binwidth = 100, fill = "skyblue", color = "black", alpha = 0.7) + 
  labs(title = "Reaction Time Overall", x = "Reaction Time (ms)", y = "Frequency") + 
  theme_minimal()

# Reaction Time 90° Histogram
ggplot(df, aes(x = ReactionTime90)) + 
  geom_histogram(binwidth = 100, fill = "skyblue", color = "black", alpha = 0.7) + 
  labs(title = "Reaction Time (90°)", x = "Reaction Time (ms)", y = "Frequency") + 
  theme_minimal()

# Reaction Time 45° Histogram
ggplot(df, aes(x = ReactionTime45)) + 
  geom_histogram(binwidth = 100, fill = "skyblue", color = "black", alpha = 0.7) + 
  labs(title = "Reaction Time (45°)", x = "Reaction Time (ms)", y = "Frequency") + 
  theme_minimal()

# Accuracy Overall Histogram
ggplot(df, aes(x = AccuracyOverall)) + 
  geom_histogram(binwidth = 8, fill = "skyblue", color = "black", alpha = 0.7) + 
  labs(title = "Accuracy Overall", x = "Percent Accuracy", y = "Frequency") + 
  theme_minimal()

# Accuracy 90° Histogram
ggplot(df, aes(x = Accuracy90)) + 
  geom_histogram(binwidth = 8, fill = "skyblue", color = "black", alpha = 0.7) + 
  labs(title = "Accuracy 90", x = "Percent Accuracy", y = "Frequency") + 
  theme_minimal()

# Accuracy 45° Histogram
ggplot(df, aes(x = Accuracy45)) + 
  geom_histogram(binwidth = 8, fill = "skyblue", color = "black", alpha = 0.7) + 
  labs(title = "Accuracy 45", x = "Percent Accuracy", y = "Frequency") + 
  theme_minimal()

# Shapiro-Wilk Test for Normality
shapiro.test(df$ReactionTimeOverall)
shapiro.test(df$ReactionTime90)
shapiro.test(df$ReactionTime45)
shapiro.test(df$AccuracyOverall)
shapiro.test(df$Accuracy90)
shapiro.test(df$Accuracy45)

# p-value < 0.05: Data are not normally distributed.
# p-value ≥ 0.05: Data are likely normally distributed

```

## Since the accuracy variables are not normally distributed, we will log transform them and then check them again. 
```{r}
# Add a small constant to handle zero values
df$AccuracyOverall_log <- log(df$AccuracyOverall + 1)
df$Accuracy90_log <- log(df$Accuracy90 + 1)
df$Accuracy45_log <- log(df$Accuracy45 + 1)

# Check normality again
shapiro.test(df$AccuracyOverall_log)
shapiro.test(df$Accuracy90_log)
shapiro.test(df$Accuracy45_log)
```
## After checking the log transformation, the overall accuracy and the accuracy for the 90 degree condition are still not normal. Thus we will proceede with non parametric tests for those variables. 


# Non-Parametric Test: 
```{r}

# Non-Parametric Test --> Mann-Whitney U Test

# For overall accuracy 
wilcox.test(AccuracyOverall_log ~ Environment, data = df)

# For accuracy of 90 degrees
wilcox.test(Accuracy90_log ~ Environment, data = df)

```
# Non Parametric Test Results: 

## Accuracy Overall: p-value = 0.04542.
### This p-value suggests a significant difference between the Urban and Rural environments in terms of the overall accuracy. 

## Accuracy 90: p-value = 1.216e-05.
### The very small p-value indicates a highly significant difference in accuracy at the 90° angle between the two environments. This result strongly suggests that there is a difference in performance between Urban and Rural environments for this specific angle.

#
#
#

# ANOVA:
```{r}

############################################################
# ANOVA for Reaction Time
# Reaction Time (45°)
anova_RT_45 <- aov(ReactionTime45 ~ Environment, data = df)

# Reaction Time (90°)
anova_RT_90 <- aov(ReactionTime90 ~ Environment, data = df)

# Overall Reaction Time (if needed)
anova_RT_overall <- aov(ReactionTimeOverall ~ Environment, data = df)

summary(anova_RT_45)
summary(anova_RT_90)
summary(anova_RT_overall)

############################################################
# ANOVA for accuracy (45 degrees only since that was the only normal one)
anova_Acc_45 <- aov(Accuracy45_log ~ Environment, data = df)
summary(anova_Acc_45)

############################################################
# ANOVA for interactions --> have to first convert it to long format in order to run the two way anova for the interactions 
df_long <- df %>%
  pivot_longer(cols = c(ReactionTime45, ReactionTime90), 
               names_to = "Angle", values_to = "ReactionTime") %>%
  mutate(Angle = ifelse(Angle == "ReactionTime45", 45, 90))  # Convert to numeric

anova_RT_interaction <- aov(ReactionTime ~ Environment * Angle, data = df_long)
summary(anova_RT_interaction)
```

# ANOVA RESULTS 
## Reaction Time ANOVA:
### - 45° Angle (Overall): Significant difference in reaction time between the environments for the 45° angle.
### - 90° Angle (Overall): Significant difference in reaction time between the environments for the 90° angle.
### - Overall Reaction Time: No significant effect of environment on overall reaction time.

## Accuracy ANOVA for 45°:
### Environment: Significant difference in accuracy between the two environments for the 45° angle.

## Interaction ANOVA: 
### The significant effect on reaction time is due to Environment, but Angle and its interaction with Environment do not seem to have a significant impact.

# Main Results: 
## Reaction Time:
### There is a strong significant difference in reaction time between Urban and Rural environments for both the 45° and 90° angles, but no significant difference for Overall Reaction Time. This suggests that the effect is specific to the angle conditions rather than the overall measure.
## Accuracy:
### There is a significant effect of environment on accuracy for the 45° angle, indicating that the environment influences how accurately participants perform at this angle.





# Post Hoc Test
```{r}

# Reaction Time (45°)
pairwise_RT_45 <- pairwise.t.test(df$ReactionTime45, df$Environment, p.adjust.method = "bonferroni")
# Reaction Time (90°)
pairwise_RT_90 <- pairwise.t.test(df$ReactionTime90, df$Environment, p.adjust.method = "bonferroni")
# Accuracy (45°)
pairwise_Acc_45 <- pairwise.t.test(df$Accuracy45_log, df$Environment, p.adjust.method = "bonferroni")


print(pairwise_RT_45)
print(pairwise_RT_90)
print(pairwise_Acc_45)

```
# Results of the Post Hoc Analysis: 
## 1. Reaction Time (45°)
### Rural vs. Urban: There is a highly significant difference in reaction time between Rural and Urban environments (p = 3.4e-09), indicating that the environment has a significant effect on reaction time at the 45° angle.

## 2. Reaction Time (90°)
### Rural vs. Urban: Similarly, there is a highly significant difference in reaction time between Rural and Urban environments (p = 8.5e-11), showing a significant effect on reaction time at the 90° angle as well.

## 3. Accuracy (45°)
### Rural vs. Urban: There is a statistically significant difference in accuracy between Rural and Urban environments (p = 0.0084), suggesting that environment has a significant effect on accuracy at the 45° angle.

## Summary:
### Reaction Time: Significant difference between Rural and Urban for both the 45° and 90° angles.
### Accuracy: Significant difference between Rural and Urban environments for the 45° angle.

# Visualizations 

```{r}
# Reaction Time (45°)
ggplot(df, aes(x = Environment, y = ReactionTime45, fill = Environment)) +
  geom_boxplot() +
  scale_fill_manual(values = c("skyblue", "lightcoral")) +  # Customize colors
  labs(title = "Reaction Time (45°) by Environment", x = "Environment", y = "Reaction Time (ms)") +
  theme_minimal()
```
#### The mean differences for reaction time of 45 degree angle between environment conditions

```{r}
# Reaction Time (90°)
ggplot(df, aes(x = Environment, y = ReactionTime90, fill = Environment)) +
  geom_boxplot() +
  scale_fill_manual(values = c("skyblue", "lightcoral")) +
  labs(title = "Reaction Time (90°) by Environment", x = "Environment", y = "Reaction Time (ms)") +
  theme_minimal()
```
#### The mean differences for reaction time of 90 degree angle between environment conditions

```{r}
# Accuracy (45°)
ggplot(df, aes(x = Environment, y = Accuracy45_log, fill = Environment)) +
  geom_boxplot() +
  scale_fill_manual(values = c("skyblue", "lightcoral")) +
  labs(title = "Accuracy (45°) by Environment", x = "Environment", y = "Accuracy (log-transformed)") +
  theme_minimal()

```
#### The mean differences for accuracy of 45 degree angle between environment conditions

```{r}
# Reaction time for both angles
df_long <- df %>%
  gather(key = "Angle", value = "ReactionTime", ReactionTime45, ReactionTime90)
# Boxplot for Reaction Times (45° and 90°) shown together 
ggplot(df_long, aes(x = Environment, y = ReactionTime, fill = Angle)) + 
  geom_boxplot(position = position_dodge(0.8)) + 
  scale_fill_manual(values = c("skyblue", "lightcoral")) +
  labs(title = "Comparison of Reaction Time (45° vs 90°) by Environment", 
       x = "Environment", y = "Reaction Time (ms)", fill = "Angle Condition") +
  theme_minimal() + 
  theme(legend.position = "top")
```
#### Combined mean reaction times for both the angles and environment

```{r}
ggplot(df, aes(x = ReactionTimeOverall, y = AccuracyOverall, color = Environment)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title = "Reaction Time vs Accuracy by Environment", 
       x = "Overall Reaction Time (ms)", y = "Accuracy (%)") + 
  theme_minimal() + 
  theme(legend.position = "top")

```
#### Scatter plot of reaction time vs accuracy for both environments with their regression lines. Both lines show that accuracy is not correlated with the reaction time in both environments.  

Additional Demographic Stats: 
```{r}

mean_age <- mean(df$Age)
print (mean_age)

```
#### The mean age for the participants was: 31.2


# Final Summary: 
##### The environment (Rural vs. Urban) significantly affects reaction times at both 45° and 90° angles, but there is no overall significant effect on reaction time when considering all angles together. The environment also has a significant effect on accuracy for the 45° angle.Angle alone doesn't seem to significantly influence reaction time in a meaningful way without the environment being considered. Additionally the visualizations show that urban conditions had a faster reaction time than rural conditions. 


